Quantitative Benchmarking for Automatically Generated Conjectures

### 2017-09-11 TEBenchmarkPaper/article.tex
Notes, and fix copypasta reference
Add table from Colton et al

% Derived from the template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments
gsave
newpath
  20 20 moveto
  20 220 lineto
  220 220 lineto
  220 20 lineto
closepath
2 setlinewidth
gsave
  .4 setgray fill
grestore
stroke
grestore
\end{filecontents*}
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
% etc.

@@

\begin{figure}
  \label{figure:list_theory}
  \begin{equation*}
    \begin{split}
      \text{Nil} : \forall (a : Type). \text{List} a \\
      \text{Cons} : \forall (a : Type). \forall (x : a). \forall (xs : \text{List} a). \text{List} a \\
      \text{head}(\text{Cons}(x, xs)) &= x \\
      \text{tail}(\text{Cons}(x, xs)) &= xs \\
      \text{append}(xs, \text{Nil}) &= xs \\
      \text{append}(\text{Cons}(x, xs), ys) &= \text{Cons}(x, \text{append}(xs, ys)) \\
      \text{reverse}(\text{Nil}) &= \text{Nil} \\
      \text{reverse}(\text{Cons}(x, xs)) &= \text{append}(\text{reverse}(xs), \text{Cons}(x, \text{Nil})) \\
      \text{length}(\text{Nil}) &= Z \\
      \text{length}(\text{Cons}(x, xs)) &= S (\text{length}(xs)) \\
      \text{map}(f, \text{Nil}) &= \text{Nil} \\
      \text{map}(f, \text{Cons}(x, xs)) &= \text{Cons}(f(x), \text{map}(f, xs)) \\
      \text{foldl}(f, x, \text{Nil}) &= x \\
      \text{foldl}(f, x, \text{Cons}(y, ys)) &= \text{foldl}(f, f(x, y), xs) \\
      \text{foldr}(f, \text{Nil}, y) &= y \\
      \text{foldr}(f, \text{Cons}(x, xs), y) &= f(x, \text{foldr}(f, xs, y))
    \end{split}
  \end{equation*}
  \caption{A simple theory defining a $\text{List}$ type and some associated
    operations. $Z$ and $S$ are from a Peano encoding of the natural numbers.
    Taken from \cite{Johansson.Dixon.Bundy:conjecture-generation}}
\end{figure}

If we have a logical theory, for example the theory of lists shown in figure
\ref{figure:list_theory}, we may want to find theorems which describe its
behaviour. For example, if this is a software library then we may want to verify
that certain (un)desirable properties do (not) hold; we may also want to
\emph{optimise} programs using this library, rewriting expressions into a form
which requires less time, memory, network usage, etc. To avoid altering the
program's result, it is important to prove particular rewrite rules correct. For
example, consider the following theorem:

@@

Discovering such theorems from an arbitrary theory requires two steps:

\begin{itemize}
\item \emph{Exploring} the theory, to find patterns which we pose as
  conjectures.
\item \emph{Proving} that those conjectures are theorems.
\end{itemize}

Lots of research has been devoted to the second step, resulting in the field of
Automated Theorem Proving (ATP) and spinoff applications (e.g. Prolog). Less
attention has been paid to the first step, that of Automated Theory Exploration.

Those ATE systems which do exist are difficult to evaluate and compare in a
quantitative way. There are three aspects to consider:

\begin{itemize}
\item The quality of each conjecture, to judge how ``interesting'' it is.
\item The \emph{set} of conjectures produced, to assess things like consistency
  and redundancy.
\item Performance of the system, to find the balance struck between output
  quality and time taken.
\end{itemize}

In particular, we must choose how to measure what is ``interesting'', and since
there is no objective answer we may find that the evaluation methods used by
each system provide results which cannot be directly compared. It is also
important to consider the goals of the system's designers, since that will
affect for which measures it performs well.

There are many aspects to consider when judging the quality of a conjecture: how
likely it is to be true, how difficult it is to prove, its implications, and so
on.

One method which is used to make these subjective, ambiguous properties more
concrete is to choose a well-studied theory, with a set of existing theorems that
are deemed to be useful or important in some way, and treat these as a
\emph{ground truth} against which we can compare conjectures generated from the
same theory.

@@

One problem with this approach is the small size of these libraries. For
example, the benchmark based on Isabelle/HOL's theory of natural numbers given
in~\cite{Johansson.Dixon.Bundy:conjecture-generation} contains
only 4 definitions and 12 theorems. Whilst such benchmarks allow objective
comparisons between different approaches, their narrow scope doesn't provide
much indication of performance in different, especially \emph{novel}, domains.

\section{Our Proposal}
\label{sec:proposal}

We propose that the extensive test suites which are already used to benchmark
automated \emph{theorem provers} can be repurposed as ground truths in the
theory \emph{exploration} setting. To do this we can take an ATP problem (for
example proving equation \ref{eq:mapreduce} from the definitions in figure
\ref{figure:list_theory}) and split apart the definitions (types and functions)
from the theorem statement itself.

The former constitute a theory, suitable for exploration by an ATE system; the
latter act as a ground truth, telling us which theorems about that theory were
``interesting'' enough to write down in a benchmark.

@@

\begin{itemize}
\item Collect together all definitions found in TIP into one large theory,
  removing $\alpha$-equivalent duplicates.
\item For each type, provide extra functions which act as constructors and
  destructors. For example, the $\text{List}$ type would have extra constructor
  functions $\text{constructor-Nil}$ and $\text{constructor-Cons}$, and
  destructor functions $\text{destructor-head}$ and $\text{destructor-tail}$.
  This allows us to ignore constructors and destructors, and concentrate solely
  on functions.
\item Collect together all theorem statements found in TIP. Update any
  references to removed definitions, and replace references to constructors with
  the associated constructor function from above.
\item \emph{Sample} a sub-set of functions from the large theory uniformly, such
  that we choose all of the names referenced by at least one of the theorems
  (for example, theorem \ref{eq:mapreduce} references the names
  $\text{map}$ and $\text{append}$).
\item Provide this sub-set as an input to the ATE system, timing how long it
  takes to complete.
\item Perform precision/recall analysis on the resulting conjectures, against
  the ground truth of TIP theorems which only reference names in this sample.
\end{itemize}

@@

\section{Discussion}
\label{sec:application}

% (e.g. philosophical bits),

>>>>

% Derived from the template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments
gsave
newpath
  20 20 moveto
  20 220 lineto
  220 220 lineto
  220 20 lineto
closepath
2 setlinewidth
gsave
  .4 setgray fill
grestore
stroke
grestore
\end{filecontents*}
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{array}
\usepackage{hhline}
\usepackage{mathtools}
\usepackage{multirow}
% etc.

@@

\subsection{Motivation}
\label{sec:motivation}

% more detail about evaluation, interestingness, etc. Reference Simon/Alan
% interestingness, e.g. examples from textbooks, etc.

% split into sub-sections, e.g. motivation, existing approaches, etc.

\begin{figure}
  \begin{equation*}
    \begin{split}
      \forall a. \text{Nil} &: \text{List} \  a \\
      \forall a. \text{Cons} &: a \rightarrow \text{List} \  a \rightarrow \text{List} \ a \\
      \text{head}(\text{Cons}(x, xs)) &= x \\
      \text{tail}(\text{Cons}(x, xs)) &= xs \\
      \text{append}(xs, \text{Nil}) &= xs \\
      \text{append}(\text{Cons}(x, xs), ys) &= \text{Cons}(x, \text{append}(xs, ys)) \\
      \text{reverse}(\text{Nil}) &= \text{Nil} \\
      \text{reverse}(\text{Cons}(x, xs)) &= \text{append}(\text{reverse}(xs), \text{Cons}(x, \text{Nil})) \\
      \text{length}(\text{Nil}) &= \text{Z} \\
      \text{length}(\text{Cons}(x, xs)) &= \text{S} (\text{length}(xs)) \\
      \text{map}(f, \text{Nil}) &= \text{Nil} \\
      \text{map}(f, \text{Cons}(x, xs)) &= \text{Cons}(f(x), \text{map}(f, xs)) \\
      \text{foldl}(f, x, \text{Nil}) &= x \\
      \text{foldl}(f, x, \text{Cons}(y, ys)) &= \text{foldl}(f, f(x, y), xs) \\
      \text{foldr}(f, \text{Nil}, y) &= y \\
      \text{foldr}(f, \text{Cons}(x, xs), y) &= f(x, \text{foldr}(f, xs, y))
    \end{split}
  \end{equation*}
  \caption{A simple theory defining a $\text{List}$ type and some associated
    operations. $\text{Z}$ and $\text{S}$ are from a Peano encoding of the natural numbers.
    Taken from \cite{Johansson.Dixon.Bundy:conjecture-generation}}
  \label{figure:list_theory}
\end{figure}

If we have a logical theory, for example the theory of lists shown in figure
\ref{figure:list_theory}, we may want to find theorems which describe its
behaviour. For a software library we may want to verify that certain
(un)desirable properties do (not) hold; we may also want to \emph{optimise}
programs using this library, rewriting expressions into a form which requires
less time, memory, network usage, etc. To avoid altering the program's result,
it is important to prove particular rewrite rules correct. Consider the
following theorem:

@@

\subsection{Theory Exploration}
\label{sec:te}

Discovering theorems from an arbitrary theory requires two steps:

\begin{itemize}
\item \emph{Exploring} the theory, to find patterns suitable for posing as
  conjectures.
\item \emph{Proving} that those conjectures are theorems.
\end{itemize}

Lots of research has been devoted to the second step, resulting in the field of
Automated Theorem Proving (ATP) and its spinoff applications. Less attention has
been paid to the first step, that of Automated Theory Exploration (ATE).

A major challenge in ATE is the requirement for conjectures to be
``interesting'', since this is an imprecise term with many different
interpretations, which is reflected in the variety of disparate approaches taken
so far.

A survey of these systems and their notions of ``interestingness'' is given by
Colton et al~\cite{colton2000notion}, for both conjectures and \emph{concepts}
(the contents of a theory, which for ATE we consider to be provided as input,
rather than part of the generating process). We summarise those aspects relevant
to conjecture formation in table \ref{table:colton}, along with later systems
we have studied.

\begin{table}
  \begin{center}
    \begin{tabular}{ |l|l|c|c|c|c|c|c| }
      \hline
      \multirow{2}{*}{\textbf{Program}}                 &
      \multirow{2}{*}{\textbf{Conjecture Types}}        &
      \multicolumn{6}{c}{\textbf{Interestingness Measures}} \\
      \hhline{~~------}
         & & \textbf{E} & \textbf{N} & \textbf{S} & \textbf{A} & \textbf{C} & \textbf{U} \\

      \hline
      AM          & if-and-only-if, implies, non-exists & X & X & X & X & X & X \\ \hline
      GT          & if-and-only-if, implies, non-exists & X & X & X & X & X & X \\ \hline
      Graffiti    & inequalities                        & X & X & X &   & X & X \\ \hline
      Bagai et al & non-exists                          &   & X &   & X & X &   \\ \hline
      HR          & if-and-only-if, implies, non-exists & X & X & X & X & X & X \\ \hline
      QuickSpec   & equalities                          & X & X &   &   & X &   \\ \hline
      IsaCoSy     & equalities                          &   & X &   &   & X &   \\ \hline
      IsaScheme   & equalities                          &   &   &   &   & X &   \\ \hline
    \end{tabular}
  \end{center}
  \caption{Classification of ATE systems from \cite{colton2000notion}, extended
    to those compared in \cite{claessen2013automating} (QuickSpec is the
    conjecture generation component of HipSpec). The interestingness measures
    are \textbf{E}mpirical plausibility, \textbf{N}ovelty,
    \textbf{S}urprisingness, \textbf{A}pplicability, \textbf{C}omprehensibility
    (low complexity) and \textbf{U}tility.}
  \label{table:colton}
\end{table}

One aspect mentioned in the survey is that ``interestingness'' is not just a
measure of output quality, but can also impact the core design of a system. For
example, the use of constraint solving in IsaCoSy is to prevent forming
conjectures which are simple substitutions of previous ones; this is because the
designers consider such conjectures to be uninteresting. This further
complicates comparison of systems based on standalone evaluations, since
assumptions about such behaviour may be implicit in the interpretation (e.g. an
analysis of IsaCoSy conjectures does not need to consider substitutions), and
each methodology will be using different assumptions.

\subsection{Existing Evaluations}
\label{sec:existing}

The variety of existing ATE systems, their different domains, and their varying
interpretations of what is interesting makes it difficult to directly compare
their performance.

Whilst all offer insights into the ATE problem, we will limit ourselves to those
operating in higher-order logic and inductive theories, since this domain is the
most relevant for our interest in analysing software. The relevant systems in
this domain are QuickSpec, operating on functions in the Haskell programming
language; and IsaCoSy and IsaScheme, which operate in the Isabelle proof
assistant.

There are three aspects to consider when we evaluate an ATE system:

\begin{itemize}
\item The quality of each conjecture, which is the ``interestingness'' discussed
  above.
\item The \emph{set} of conjectures produced, to assess things like consistency
  and redundancy.
\item Performance of the system, to find the balance struck between output
  quality and time taken.
\end{itemize}

Existing evaluation of these three systems have assessed the interestingness of
their output by operating in well-studied theories, and taking a set of existing
theorems as a \emph{ground truth} to compare against.

@@

One problem with this choice is the small size of these libraries. For example,
the benchmark based on Isabelle/HOL's theory of natural numbers given
in~\cite{Johansson.Dixon.Bundy:conjecture-generation} contains only 4
definitions and a ground truth of 12 theorems. Whilst such benchmarks allow
objective comparisons between different approaches, their narrow scope doesn't
provide much indication of performance in different, especially \emph{novel}, domains.

\section{Theory Exploration Benchmark}
\label{sec:proposal}

We propose a larger, more ambitious method for evaluating ATE systems, which we
call the Theory Exploration Benchmark (TEB). We take a similar approach to prior
evaluations, but our ground truth comes from the extensive test suites which are
already used to benchmark automated \emph{theorem provers}; which we repurpose
for use in the theory \emph{exploration} setting.

To do this we can take a set of ATP tasks, for example proving the equation
\ref{eq:mapreduce} from the theory in figure \ref{figure:list_theory}, and we
split apart the definitions (to form a theory) from the theorem statements (to
form our ground truth).

@@

% Better presentation, flowchart?
\begin{itemize}
\item Collect together all definitions found in TIP into one large theory,
  removing $\alpha$-equivalent duplicates.
\item For each type, provide extra functions which act as constructors and
  destructors. For example, the $\text{List}$ type would have extra constructor
  functions $\text{constructor-Nil}$ and $\text{constructor-Cons}$, and
  destructor functions $\text{destructor-head}$ and $\text{destructor-tail}$.
  This allows us to ignore constructors and destructors, and concentrate solely
  on functions.
\item Collect together all theorem statements found in TIP. Update any
  references to removed definitions, and replace references to constructors with
  the associated constructor function from above.
\item \emph{Sample} a sub-set of functions from the large theory uniformly, such
  that we choose all of the names referenced by at least one of the theorems
  (for example, theorem \ref{eq:mapreduce} references the names
  $\text{map}$ and $\text{append}$).
\item Provide this sub-set as an input to the ATE system, timing how long it
  takes to complete.
\item Perform precision/recall analysis on the resulting conjectures, against
  the ground truth of TIP theorems which only reference names in this sample.
\end{itemize}

@@

\section{Discussion}
\label{sec:discussion}

% philosophical bits
